---
title: "H-1B Data Analysis Dashboard (FY24)"
author: "Geethakrishna Puligundla"
date: "04/29/2025"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: journal
runtime: shiny
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = FALSE,       # Hide code chunks in the dashboard output
  warning = FALSE,    # Suppress warnings
  message = FALSE     # Suppress messages
)

library(flexdashboard) # For creating the dashboard layout
library(shiny)         # For interactive elements (inputs, outputs)
library(tidyverse)     # For data manipulation (dplyr, ggplot2, etc.)
library(plotly)        # For interactive plots
library(DT)            # For interactive tables
library(scales)        # For formatting plot axes (e.g., comma)

# Read raw data
h1b_data_raw <- read.csv("Employer_Information_F24.csv")

# Data Cleaning and Transformation Pipeline
h1b_data <- h1b_data_raw %>%
  # Select and rename relevant columns
  select(
    Employer_Name = Employer..Petitioner..Name,
    Tax_ID = Tax.ID,
    NACIS_CODE = Industry..NAICS..Code,
    Petitioner_City = Petitioner.City,
    Petitioner_State = Petitioner.State,
    Petitioner_Zip = Petitioner.Zip.Code,
    Initial_Approval = Initial.Approval,
    Initial_Denial = Initial.Denial,
    Continuing_Approval = Continuing.Approval,
    Continuing_Denial = Continuing.Denial
  ) %>%
  
  # Handle missing values and data types
  mutate(
    # Replace empty strings in Employer_Name with NA
    Employer_Name = na_if(Employer_Name, ""),
    Initial_Approval = coalesce(as.integer(Initial_Approval), 0L),
    Initial_Denial = coalesce(as.integer(Initial_Denial), 0L),
    Continuing_Approval = coalesce(as.integer(Continuing_Approval), 0L),
    Continuing_Denial = coalesce(as.integer(Continuing_Denial), 0L)
  ) %>%
  
  # Filter out rows with essential missing data or no activity
  filter(
    !is.na(Employer_Name),
    !is.na(Tax_ID),
    !is.na(Petitioner_Zip),
    !(Initial_Approval == 0 & Initial_Denial == 0 &
      Continuing_Approval == 0 & Continuing_Denial == 0)
  ) %>%
  
  # Aggregate data by Employer_Name to handle potential duplicates
  group_by(Employer_Name) %>%
  summarize(
    # Take the first non-missing value for identifiers within each group
    Tax_ID = first(na.omit(Tax_ID)),
    NACIS_CODE = first(na.omit(NACIS_CODE)),
    Petitioner_City = first(na.omit(Petitioner_City)),
    Petitioner_State = first(na.omit(Petitioner_State)),
    Petitioner_Zip = first(na.omit(Petitioner_Zip)),
  
    Initial_Approval = sum(Initial_Approval, na.rm = TRUE),
    Initial_Denial = sum(Initial_Denial, na.rm = TRUE),
    Continuing_Approval = sum(Continuing_Approval, na.rm = TRUE),
    Continuing_Denial = sum(Continuing_Denial, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  
  # Create calculated fields for analysis and filtering
  mutate(
    Total_H1B = Initial_Approval + Initial_Denial + Continuing_Approval + Continuing_Denial,
    Total_Lottery_H1B = Initial_Approval + Initial_Denial,
    Total_Other = Continuing_Approval + Continuing_Denial
  ) %>%
  
  # Filter out rows where Petitioner_State is NA, as it's used in the selector
  filter(!is.na(Petitioner_State))

```

# Input Choices

```{r}

# Get unique states for the dropdown selector
unique_states <- sort(unique(h1b_data$Petitioner_State))
# Create named list for choices, including an "All States" option
state_choices <- c("All States" = "All", unique_states)

max_lottery_petitions <- max(1, h1b_data$Total_Lottery_H1B, na.rm = TRUE)

```

Column {.sidebar data-width=300}
-----------------------------------------------------------------------

### Controls

This sidebar contains controls to filter the H-1B data displayed in the main panels.

```{r inputs}
# --- Input Widgets ---

# Input 1: Dropdown for selecting Petitioner State
selectInput(
  inputId = "state_select",           # ID to reference this input in server logic
  label = "Select State:",            # Label displayed above the dropdown
  choices = state_choices,            # List of choices generated in setup
  selected = "All"                    # Default selected value
)

# Input 2: Slider for filtering by minimum initial petitions
sliderInput(
  inputId = "min_lottery_h1b",        # ID to reference this input
  label = "Minimum Initial Petitions:",# Label displayed above the slider
  min = 0,                            # Minimum value of the slider
  max = max_lottery_petitions,        # Maximum value (calculated in setup)
  value = 1,                          # Default starting value
  step = 1                            # Increment step size
)

```

Column {data-width=700}
-----------------------------------------------------------------------

### Top Employers by Initial Petitions

```{r top_employers_plot}
# Create a reactive expression that filters the data based on user inputs
filtered_data <- reactive({
  data <- h1b_data

  # Apply state filter if a specific state is selected
  if (input$state_select != "All") {
    data <- data %>% filter(Petitioner_State == input$state_select)
  }

  # Apply minimum initial petitions filter using the slider value
  data <- data %>% filter(Total_Lottery_H1B >= input$min_lottery_h1b)

  data
})


renderPlotly({

  # Prepare data specifically for the plot (Top N employers)
  plot_data <- filtered_data() %>%
    arrange(desc(Total_Lottery_H1B)) %>%
    head(10)

  # Check if plot_data has any rows after filtering
  validate(
    need(nrow(plot_data) > 0, "No employers match the selected criteria.")
  )

  # Create the ggplot object
  p <- ggplot(plot_data, aes(
                x = reorder(Employer_Name, Total_Lottery_H1B),
                y = Total_Lottery_H1B,
                # Define custom hover text for plotly
                text = paste("Employer:", Employer_Name,
                             "<br>State:", Petitioner_State, # Added State info
                             "<br>Initial Petitions:", Total_Lottery_H1B)
              )) +
    geom_col(fill = "steelblue", alpha = 0.8) + # Bar chart layer
    coord_flip() + 
    labs(
      title = paste("Top 10 Employers by Initial Petitions",
                    ifelse(input$state_select == "All", "(All States)", paste("in", input$state_select))),
      subtitle = paste("Minimum Initial Petitions:", input$min_lottery_h1b),
      x = "Employer Name",
      y = "Number of Initial Petitions (Approved + Denied)"
    ) +
    scale_y_continuous(labels = scales::comma) +
    theme_minimal(base_size = 11) + 
    theme(plot.title = element_text(size = 14, face = "bold"),
          plot.subtitle = element_text(size = 10))

  # Convert the ggplot object to an interactive plotly object
  ggplotly(p, tooltip = "text") %>%
    layout(title = list(text = paste("Top 10 Employers by Initial Petitions",
                                      ifelse(input$state_select == "All", "(All States)", paste("in", input$state_select)),
                                      '<br>',
                                      '<sup>',paste("Minimum Initial Petitions:", input$min_lottery_h1b),'</sup>'),
                      font = list(size = 14)))

})

```

Column {data-width=500}
-----------------------------------------------------------------------

### Filtered Employer Data Table

```{r filtered_data_table}
# Use renderDataTable to create an interactive table output
renderDataTable({

  # Access the reactive filtered data
  data_for_table <- filtered_data() %>%
    select(
      Employer_Name,
      Petitioner_State,
      Petitioner_City,
      Total_Lottery_H1B,
      Initial_Approval,
      Initial_Denial
    ) %>%
    arrange(desc(Total_Lottery_H1B))

   validate(
    need(nrow(data_for_table) > 0, "No data available for the selected filters.")
  )

  # Create the DataTable object
  datatable(
    data_for_table,
    rownames = FALSE,           # Don't show row numbers
    filter = 'top',             # Add column filters at the top
    extensions = c('Buttons'), # Add extensions for extra features
    options = list(
      pageLength = 25,          # Default number of rows per page
      scrollX = TRUE,           # Enable horizontal scrolling for wide tables
      scroller = TRUE,           # Use Scroller extension for large tables (performance)
      dom = 'Bfrtip'     
    ),
    caption = htmltools::tags$caption(
      style = 'caption-side: top; text-align: left;', 
      'Filtered H-1B Employer Data (FY2024)'
    )
  )
})

